Assessing Bridge Condition in New York State Using Multiple Linear Regression

### 2024-12-04 paper/paper.qmd
Updated discussion

```
@fig-age-histogram shows distribution of bridge ages at the time of inspection, overlaid with red dashed lines and a blue density curve. The red dashed lines mark intervals of 25 years and align with clear peaks in the histogram, suggesting periodic patterns in bridge construction or major renovations. Prominent clusters are observed around 25, 50, and 75 years of age, indicating that many bridges were likely built or reconstructed during these timeframes. The blue density curve provides a smooth visualization of the distribution, confirming the periodic modes while highlighting the overall trends. Beyond 100 years, the frequency drops significantly, with very few bridges older than this still in operation, reflecting the typical lifecycle of infrastructure. The concentration of bridges between 20 and 80 years old suggests that a significant portion of bridges are mid-life and may soon require maintenance or replacement.It highlights both the historical patterns of bridge construction and the current age distribution of bridges in the dataset.
```{r}
#| label: fig-ageatinspection-condition
#| fig-cap: "Relationship Between Bridge Age at Inspection and Bridge Condition Rating. Each blue dot represents a bridge, showing its age at the time of inspection and the corresponding condition rating obtained from that inspection."
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

## Age Distribution Patterns {#sec-age-distribution}
The distribution of bridge ages at inspection demonstrates distinct peaks approximately every 25 years, reflecting historical trends in bridge construction or renovation efforts. These periodic patterns likely correspond to infrastructure development initiatives or funding cycles at specific points in time. Understanding these trends is crucial for proactive resource allocation, as large clusters of aging bridges may require simultaneous maintenance or replacement. Planning for these future needs will help mitigate risks associated with concentrated structural deterioration.

@@

## Model Limitations and Future Improvements
The model's low \( R^2 \) value, 0.396, suggests it does not fully capture the range of factors influencing bridge conditions. Key omitted variables, such as traffic volume, environmental stressors, and material quality, may significantly impact bridge deterioration. Future models should incorporate these variables, which can be obtained from traffic monitoring systems, meteorological data, and engineering databases, to enhance predictive accuracy. Furthermore, the assumption of constant variance is invalid for bridges with a condition rating below 4.5, indicating the current model is unsuitable for severely damaged bridges. While these bridges represent a small proportion of the dataset, their exclusion limits the model's scope. Future research should develop a separate model for low-condition bridges to address their unique deterioration dynamics.

>>>>

```
@fig-age-histogram shows distribution of bridge ages at the time of inspection, overlaid with red dashed lines and a blue density curve. The red dashed lines mark intervals of 25 years and align with clear peaks in the histogram, suggesting periodic patterns in bridge construction or major renovations. Prominent clusters are observed around 25, 50, and 75 years of age, indicating that many bridges were likely built or reconstructed during these timeframes. The blue density curve provides a smooth visualization of the distribution, confirming the periodic modes while highlighting the overall trends. This point will be further discussion in @sec-age-distribution. Beyond 100 years, the frequency drops significantly, with very few bridges older than this still in operation, reflecting the typical lifecycle of infrastructure. The concentration of bridges between 20 and 80 years old suggests that a significant portion of bridges are mid-life and may soon require maintenance or replacement.It highlights both the historical patterns of bridge construction and the current age distribution of bridges in the dataset.
```{r}
#| label: fig-ageatinspection-condition
#| fig-cap: "Relationship Between Bridge Age at Inspection and Bridge Condition Rating. Each blue dot represents a bridge, showing its age at the time of inspection and the corresponding condition rating obtained from that inspection."
#| echo: false
#| warning: false
#| message: false
#| fig-pos: "H"

@@

## Age Distribution Patterns {#sec-age-distribution}
The distribution of bridge ages at inspection reveals distinct peaks approximately every 25 years, indicating a recurring cycle of bridge construction and major renovation. This pattern suggests that every 25 years, a significant number of bridges reach the point where extensive maintenance or replacement is necessary. The previous peak occurred roughly 25 years ago, meaning the next wave of large-scale repairs or new construction is approaching. Recognizing this pattern highlights the importance of early planning and budgeting to address the anticipated surge in infrastructure needs effectively, ensuring resources are ready to manage the simultaneous demands of aging bridges.

@@

## Model Limitations and Future Improvements
The model's low $R^2$ value, 0.396, suggests it does not fully capture the range of factors influencing bridge conditions. Key omitted variables, such as traffic volume, environmental stressors, and material quality, may significantly impact bridge deterioration. Future models should incorporate these variables, which can be obtained from traffic monitoring systems, meteorological data, and engineering databases, to enhance predictive accuracy. Furthermore, the assumption of constant variance is invalid for bridges with a condition rating below 4.5, indicating the current model is unsuitable for severely damaged bridges. While these bridges represent a small proportion of the dataset, their exclusion limits the model's scope. Future research should develop a separate model for low-condition bridges to address their unique deterioration dynamics.
